Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2021-12-29 analysis/paper/paper.Rmd
updates to management implications

At least five of the eight populations in the Upper Salmon MPG must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for the recovery of the Snake River Evolutionary Significant Unit. Populations with the ESU have substantial cultural value, support downriver mainstem Snake and Columbia River commercial and subsistence fisheries, and support local fisheries and economies in years with sufficient abundance.

We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces annual adult returns by ***xxx - x,xxx*** (perhaps express as percent, too) to upriver populations. Presumably, that reduction in adult returns impacts both the ESA-listed natural populations in the Upper Salmon River MPG and two hatchery populations in the Upper Salmon, Pahsimeroi and Sawtooth hatcheries, which provide for recrational fishing opportunities. Consequently, reducing juvenile Chinook Salmon predation mortalities at Deadwater Slough can potentially benefit multiple upriver natural and hatchery populations upriver. Because the Deadwater Slough is part of the migratory pathway for emigrating salmonids, including for multiple species and populations, the impact of that benefit would likely be higher than tributary habitat rehabilitation actions, which typically benefit a single population. Moreover, the deepened, slack water conditions that favor northern pikeminnow at Deadwater Slough are indirectly the result of manmade activities i.e., the failure of a manmade mining reservoir dam. Given these reasons, it seems that Deadwater Slough could be a candidate for management or restoration actions to benefit local Chinook Salmon populations.

>>>>

<!-- Again, some of the following numbers need to be updated after results are formalized -->
Chinook Salmon populations in the Upper Salmon River MPG support local fisheries and economies in years with sufficient abundance and have substantial cultural value in the region. These populations additionally support downriver commercial and subsistence fisheries in the Snake and Columbia rivers. Recovery of the MPG is desired which requires that at least five of the eight populations must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. Spawner abundance is perhaps the most important metric considered in determining a population's viability. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces annual adult natural-origin returns by 26.4 to 145.6 to upriver populations. Although this does not sound like a substantial amount, it represents 4.7% - 26.1% of the total mean adult escapement of 558 reported for the Upper Salmon River MPG in 2017 - 2019 [@Kinzer2020]. Presumably, reductions in adult returns also affect natural-origin Sockeye Salmon and steelhead as well as hatchery populations in the Upper Salmon which provide for recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations for multiple ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for emigrating salmonids, including for multiple species and populations, the impact of that benefit could be higher than tributary rehabilitation actions, which typically benefit a single population. Moreover, the deepened, slack water conditions that favor northern pikeminnow at Deadwater Slough are indirectly the result of manmade activities i.e., the failure of a manmade mining reservoir dam. Given these reasons, it seems that Deadwater Slough could be a candidate for management or restoration actions to benefit local Chinook Salmon populations.
